Plug-and-play dual-tree algorithm runtime analysis

### 2014-11-12 bounds-files/general-runtime-bound.tex
Remove negative vspace and remove NIPS formatting.
Update; rewritten intro, imbalance, and bound.

\begin{thm}
\label{thm:ct-runtime}
%For any dual-tree algorithm using cover trees and the standard cover tree
%dual-tree pruning traversal (Algorithm \ref{alg:cover-tree-dual}) with reference
%set $S_r$ that has expansion constant $c_r$ and query set $S_q$, the runtime is
%bounded above by $O(c_r^4 | R^* | \nu \chi \psi N)$, where $ | R^* | $ is
%the maximum size of the reference set $R$ (line
%\ref{alg:line:ct-dual-input}) at any point in the dual-tree pruning traversal,
%$\chi$ is the longest possible running time of \texttt{BaseCase()}, and $\psi$
%is the longest possible running time of \texttt{Score()}.
Given a reference set $S_r$ of size $N$ with an expansion constant $c_r$ and a
set of queries $S_q$ of size $O(N)$, a standard cover tree based dual-tree
algorithm (Algorithm \ref{alg:cover-tree-dual}) takes $O(c_r^4 | R^* | \nu \chi
\psi N)$, where $ | R^* | $ is the maximum size of the reference set $R$ (line
\ref{alg:line:ct-dual-input}) during the dual-tree recursion, $\chi$ is
the maximum possible runtime of \texttt{BaseCase()}, and $\psi$ is the maximum
possible runtime of \texttt{Score()}.
\end{thm}

\vspace*{-1.23em}
\begin{proof}
Consider a reference recursion (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end}).
The work done in the base case loop from lines
\ref{alg:line:ct-dual-base-case-start}--\ref{alg:line:ct-dual-base-case-end} is
$O(\chi | R |)$.  Define $| R^* |$ to be the largest set $|R|$ for any scale
$s_r^{\max}$ and any query node $\mathscr{N}_q$ during the course of the
algorithm; then, it is true that $| R | \le | R^* |$.
%
Then, lines \ref{alg:line:ct-dual-ref-children} and
\ref{alg:line:ct-dual-ref-score} take $O(c_r^4 \psi | R |) \le O(c_r^4 \psi |
R^* |)$ time, because each reference node has $c_r^4$ children.  So, one
full reference recursion takes $O(c_r^4 \psi | R^* |)$ time.

Now, note that there are $O(N)$ nodes in $\mathscr{T}_q$.  Thus, line
\ref{alg:line:ct-dual-query-recursion} is visited $O(N)$ times.  Each of these
$O(N)$ visits to line \ref{alg:line:ct-dual-query-recursion} implies a
recursion, in which the reference set is descended up to $\nu$ times (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
before the query node is descended or the algorithm terminates.  In addition,
each $O(N)$ recursion implies an $O(\psi |R|) \le O(\psi |R^*|)$ operation for
the calculation of $R'$ (line \ref{alg:line:ct-dual-query-pruning}).  Thus, the
full runtime of the algorithm is bounded as $O(c_r^4 |R^*| \nu \chi \psi N +
\psi |R^*| N) = O(c_r^4 |R^*| \nu \chi \psi N)$.
\end{proof}
\vspace*{-0.8em}

>>>>

Perhaps more interesting than measures of tree imbalance is the way cover trees
are actually used in dual-tree algorithms.  Although cover trees were originally
intended for nearest neighbor search (see Algorithm \texttt{Find-All-Nearest},
\cite{langford2006}), they can be adapted to a wide variety of problems: minimum
spanning tree calculation \cite{march2010fast}, approximate nearest neighbor
search \cite{ram2010rank}, Gaussian processes posterior calculation
\cite{moore2014fast}, max-kernel search \cite{curtin2013fast,
curtin2014dual} are some examples.  Further, through the tree-independent
dual-tree algorithm abstraction of Curtin et~al. \cite{curtin2013tree}, other
existing dual-tree algorithms can easily be adapted for use with cover trees.

In the framework of tree-independent dual-tree algorithms
\cite{curtin2013tree}, all that is necessary to describe a dual-tree algorithm
is a point-to-point base case function (\texttt{BaseCase()}) and a node-to-node
pruning rule (\texttt{Score()}).  These functions, which are often very
straightforward, are then paired with a type of tree and a pruning dual-tree
traversal to produce a working algorithm.  In later sections, we will consider
specific examples.

When using cover trees, the typical pruning dual-tree traversal is an adapted
form of the original nearest neighbor search algorithm
(\texttt{Find-All-Nearest}, \cite{langford2006}); this traversal is implemented
in both the cover tree reference implementation and in the more flexible {\bf
mlpack} library \cite{curtin2013mlpack}.  The algorithm is given in Algorithm
\ref{alg:cover-tree-dual} and was originally presented by Curtin and Ram
\cite{curtin2014dual}.  Initially, it is called with the root of the query tree
and a reference set containing only the root of the reference tree.

\begin{algorithm}[htb]
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, set of reference nodes
$R$ \label{alg:line:ct-dual-input}
    \STATE {\bfseries Output:} none

    \medskip
    \STATE $s^{\max}_r \gets \max_{\mathscr{N}_r \in R} s_r$
    \IF{$(s_q < s^{\max}_r)$ {\bf or} $(s^{\max}_r = -\infty)$}
\label{alg:line:ct-dual-ref-recursion-start}
      \FORALL{$\mathscr{N}_r \in R$} \label{alg:line:ct-dual-base-case-start}
        \STATE \texttt{BaseCase($p_q$, $p_r$)}
      \ENDFOR \label{alg:line:ct-dual-base-case-end}
      \STATE $R_r \gets \{ \mathscr{N}_r \in R : s_r = s^{\max}_r \}$
\label{alg:line:ct-dual-ref-set}
      \STATE $R_{r - 1} \gets \{ \mathscr{C}(\mathscr{N}_r) : \mathscr{N}_r \in
R_r \} \cup (R \setminus R_r)$ \label{alg:line:ct-dual-ref-children}
      \STATE $R'_{r - 1} \gets \{ \mathscr{N}_r \in R_{r - 1} :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-ref-score}
      \STATE recurse with $\mathscr{N}_q$ and $R'_{r - 1}$
\label{alg:line:ct-dual-ref-recursion-end}
    \ELSE \label{alg:line:ct-dual-query-recursion-start}
      \FORALL{$\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q)$}
        \STATE $R' \gets \{ \mathscr{N}_r \in R :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-query-pruning}
        \STATE recurse with $\mathscr{N}_{qc}$ and $R'$
\label{alg:line:ct-dual-query-recursion}
      \ENDFOR \label{alg:line:ct-dual-query-recursion-end}
    \ENDIF
  \end{algorithmic}
  \caption{The standard pruning dual-tree traversal for cover trees.}
  \label{alg:cover-tree-dual}
\end{algorithm}

This dual-tree recursion is a depth-first recursion in the query tree and a
breadth-first recursion in the reference tree.  It descends either the query
tree or the reference tree (not both) during a single iteration, and the
conditional in line \ref{alg:line:ct-dual-ref-recursion-start} ensures that the
combinations of query and reference nodes that are being visited are close in
scale.

Due to the favorable theoretical properties of the cover tree, we can prove
general runtime bounds when this traversal is used.

\begin{thm}
\label{thm:ct-runtime}
%For any dual-tree algorithm using cover trees and the standard cover tree
%dual-tree pruning traversal (Algorithm \ref{alg:cover-tree-dual}) with reference
%set $S_r$ that has expansion constant $c_r$ and query set $S_q$, the runtime is
%bounded above by $O(c_r^4 | R^* | \nu \chi \psi N)$, where $ | R^* | $ is
%the maximum size of the reference set $R$ (line
%\ref{alg:line:ct-dual-input}) at any point in the dual-tree pruning traversal,
%$\chi$ is the longest possible running time of \texttt{BaseCase()}, and $\psi$
%is the longest possible running time of \texttt{Score()}.
Given a reference set $S_r$ of size $N$ with an expansion constant $c_r$ and a
set of queries $S_q$ of size $O(N)$, a standard cover tree based dual-tree
algorithm (Algorithm \ref{alg:cover-tree-dual}) takes $O(c_r^4 | R^* | \chi
\psi (N + \theta))$, where $ | R^* | $ is the maximum size of the reference set $R$ (line
\ref{alg:line:ct-dual-input}) during the dual-tree recursion, $\chi$ is
the maximum possible runtime of \texttt{BaseCase()}, $\psi$ is the maximum
possible runtime of \texttt{Score()}, and $\theta$ is a measure of imbalance
defined by

\begin{equation}
\theta = \max(\log_2(\eta_r / \eta_q), 0) + i(\mathscr{T}_q) + N
\max(\log_2(\delta_q / \delta_r), 0)
\end{equation}

\noindent with $\eta_q$ as the maximum pairwise distance between points in
$S_q$, $\delta_q$ as the minimum nonzero pairwise distance between points in
$S_q$, $\eta_r$ as the maximum pairwise distance between points in $S_r$, and
$\delta_r$ as the minimum nonzero pairwise distance between points in $S_r$.
\end{thm}

\begin{proof}
First, split the algorithm into two parts: reference recursions (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
and query recursions (lines
\ref{alg:line:ct-dual-query-recursion-start}--\ref{alg:line:ct-dual-query-recursion-end}).
The runtime of the algorithm is bounded as the runtime of a reference recursion
times the total number of reference recursions plus the total runtime of all
query recursions.

Consider a reference recursion (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end}).
The work done in the base case loop from lines
\ref{alg:line:ct-dual-base-case-start}--\ref{alg:line:ct-dual-base-case-end} is
$O(\chi | R |)$.  Define $| R^* |$ to be the largest set $|R|$ for any scale
$s_r^{\max}$ and any query node $\mathscr{N}_q$ during the course of the
algorithm; then, it is true that $| R | \le | R^* |$.
%
Then, lines \ref{alg:line:ct-dual-ref-children} and
\ref{alg:line:ct-dual-ref-score} take $O(c_r^4 \psi | R |) \le O(c_r^4 \psi |
R^* |)$ time, because each reference node has $c_r^4$ children.  So, one
full reference recursion takes $O(c_r^4 \psi \chi | R^* |)$ time.

Now, note that there are $O(N)$ nodes in $\mathscr{T}_q$.  Thus, line
\ref{alg:line:ct-dual-query-recursion} is visited $O(N)$ times.  The amount of
work in line \ref{alg:line:ct-dual-query-pruning}, like in the reference
recursion, is bounded as $O(c_r^4 \psi | R^* |)$. Therefore, the total
runtime of all query recursions is $O(c_r^4 \psi | R^* | N)$.

Lastly, we must bound the total number of reference recursions.  Reference
recursions happen in three cases: \textit{(1)} $s_r^{\max}$ is greater than the
scale of the root of the query tree (no query recursions have happened yet);
\textit{(2)} $s_r^{\max}$ is less than or equal to the scale of the root of the
query tree, but is greater than the minimum scale of the query tree that is not
$-\infty$; \textit{(3)} $s_r^{\max}$ is less than the minimum scale of the query
tree that is not $-\infty$.

First, consider case \textit{(1)}.  If the scale of the root of the query tree
is $s_q^T$ and the scale of the root of the reference tree is $s_r^T$, then
$s_r^T - s_q^T$ recursions of this sort occur before the first query recursion.

Next, consider case \textit{(2)}.  Each query recursion implies at least one
reference recursion; thus there will be at least $O(N)$ recursions.  Any
extra recursions above and beyond this occur because the query node does not
have scale which is exactly one less than its parent.  When the query node is
not a leaf, the number of extra reference recursions is bounded by the
difference between the scales of the query node and its parent minus one (one is
subtracted to account for the one reference recursion that must happen).  Each
query node is visited once, and therefore the total number of extra recursions
when the query node is not a leaf is bounded by

\begin{equation}
\sum_{\mathscr{N}_q \in \mathscr{T}_q} \sum_{\mathscr{N}_{qc} \in
\mathscr{C}(\mathscr{N}_q), s_{qc} \ne -\infty} s_q - s_{qc} - 1.
\end{equation}

When the query node is a leaf, then the number of recursions in case
\textit{(2)} for that query node is bounded by the difference between the scale
of the query node's parent and the smallest scale of a non-leaf node in the
query tree (call this $s_q^{\min}$) minus one:

\begin{equation}
\sum_{\mathscr{N}_q \in \mathscr{T}_q} \sum_{\mathscr{N}_{qc} \in
\mathscr{C}(\mathscr{N}_q), s_{qc} = -\infty} \max(s_q - s_q^{\min} - 1, 0).
\end{equation}

The $\max$ is necessary to account for the case where the query node's parent
has scale $s_q^{\min}$; no reference recursions of type \textit{(2)} happen at
all in this case.  Therefore the total number of extra reference recursions of
type \textit{(2)} is bounded by $b$, where

\begin{equation}
b = \sum_{\mathscr{N}_q \in \mathscr{T}_q} \left(\sum_{\mathscr{N}_{qc} \in
\mathscr{C}(\mathscr{N}_q), s_{qc} \ne -\infty} s_q - s_{qc} - 1 \right) +
\left(\sum_{\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q), s_{qc} = -\infty}
\max(s_q - s_q^{\min} - 1, 0) \right).
\end{equation}

(Note that $b$ can be calculated in a single $O(N)$ traversal of the tree.)

Lastly, case \textit{(3)} occurs when the reference scale is less than the
minimum query scale $s_q^{\min}$.  Define the minimum reference scale as
$s_r^{\min}$.  If $s_q^{\min} - s_r^{\min} > 0$, each query node may have up to
$s_q^{\min} - s_r^{\min}$ reference recursions of type \textit{(3)}, and there
are $N$ query nodes.  Thus, the total number of reference recursions of type
\textit{(3)} is bounded by $N \max(s_q^{\min} - s_r^{\min}, 0)$.

Combining these three cases, we get the following expression for an upper bound
$\theta$ on the total number of reference recursions:

\begin{equation}
\theta = \max(s_r^T - s_q^T, 0) + b + N (s_q^{\min} - s_r^{\min}).
\end{equation}

Now, we can express these in terms of more common constants which are related to
the aspect ratio of the data.  Define $\delta_q$ as the minimum nonzero pairwise
distance in $S_q$ and $\eta_q$ as the furthest pairwise distance in $S_q$ (and
define $\delta_r$ and $\eta_r$ accordingly for the set $S_r$).  Then, $s_r^T =
\lceil \log_2 \eta_r \rceil$, $s_q^T = \lceil \log_2 \eta_q$, $s_r^{\min} =
\lceil \log_2 \delta_r \rceil$, and $s_q^{\min} = \lceil \log_2 \delta_q
\rceil$.  This gives a more reasonable expression:

\begin{equation}
\theta = \max(\log_2(\eta_r / \eta_q), 0) + b + N \max(\log_2(\delta_q /
\delta_r), 0).
\end{equation}

Returning to our original line of reasoning, we can combine these results to
give the result of the theorem:

\begin{equation}
O(c_r^4 \psi \chi | R^* | \theta) + O(c_r^4 \psi | R^* | N) \sim O(c_r^4 \psi
\chi |R^*| (N + \theta).
\end{equation}
\end{proof}

A quick discussion of the numerous constants we have introduced and what they
mean.  In the monochromatic case, the terms of $\theta$ depending on $\eta_q$,
$\eta_r$, $\delta_q$, and $\delta_r$ disappear entirely.  However we are left
with the $b$ term, which is kind of a measure of imbalance of the tree.  The
best case is a case where the tree is perfectly balanced and there are no
`missing levels' where a child of a node is more than one level away from its
parent.  In practice, this quantity is probably much less than $N$ and should
not scale significantly with $N$ (those points which contribute to $b$ the most
will probably be outliers).  In the worst case, $b$ is bounded by the maximum
depth of the tree multiplied by the number of points, which is trivially $O(N
\log (\eta_q / \eta_r)$.  It is likely that a tighter bound of $O(N \log N)$
could be established.

There are two corner cases for the non-monochromatic case: all extra reference
recursions are of type \textit{(1)}, and all extra reference recursions are of
type \textit{(3)}.  The first case presents the best situation, because all of
the reference recursions are done before the first query node is recursed in.
Therefore the bulk of the reference recursion work is done by with the query
tree root as the query node; this is the best case.  The worst case is the
latter, where recursion in the reference tree cannot begin until reaching the
bottom of the query tree.  Thus, no pruning can occur until reaching the leaves
of the query tree, and the recursion is essentially equivalent to a single-tree
recursion for each query point plus the added overhead of the recursion of the
query tree with one reference node.  This second case degenerates to something
like $O(N \log N)$ or $O(N^2)$ time depending on how you look at it.

### 2014-11-12 bounds-files/imbalance.tex
I'll figure out how to do the bound soon.
Update; rewritten intro, imbalance, and bound.

>>>>

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $i(\mathscr{T}_b) = 0$.  A worst-case cover tree $\mathscr{T}_w$ which
is entirely a chain-like structure with maximum scale $s_{\max}$ and minimum
scale $s_{\min}$ will have imbalance $i(\mathscr{T}_w) = N (s_{\max} -
s_{\min})$.  Because of this chain-like structure, each level has only one node
and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
that the imbalance is actually quadratic in $N$!

As we will see later empirically, the notion of cover tree imbalance does a good
job of capturing the `goodness' of a tree.  We will use this notion in order to
quantify the worst-case performance of dual-tree algorithms using the cover
tree.

{\bf TODO: Can any bound be proven on $i(\mathscr{T})$?  I haven't thought of
anything successfully yet.}

### 2014-11-12 bounds-files/sec01-supporting-lemmas.tex
Remove negative vspace and remove NIPS formatting.
Update; rewritten intro, imbalance, and bound.

As we will be making heavy use of trees, we must establish notation (taken from
\cite{curtin2013tree}).  $\mathscr{T}_i$ represents a tree; $\mathscr{N}_i$
represents a node; $\mathscr{C}(\mathscr{N}_i)$ is the set of children of the
node $\mathscr{N}_i$; $\mathscr{P}(\mathscr{N}_i)$ is the set of points held in
a node; and $\mathscr{D}^p(\mathscr{N}_i)$ is the set of all descendant points
of the node $\mathscr{N}_i$.

First, we will introduce the cover tree\footnote{In our discussions, we will
only consider the explicit cover tree, as opposed to the implicit cover tree
\cite{langford2006}.  This has the effect of making the traversal more complex,
but it helps simplify our proofs.  Familiarity with the implicit cover tree
representation should not
be required for understanding any of our work here.}.  The cover tree is a
leveled hierarchical data structure originally proposed for the task of nearest
neighbor search by Beygelzimer et~al.~\cite{langford2006}.  Each cover tree node
holds one point $p_i$ and may have a number of child nodes; if the cover tree
node has children, then one child will be the self-child, which holds the same
point.  In addition, each level in the tree is indexed by an integer scale
$s_i$; as the tree is descended, $s_i$ decreases.  Every descendant of a node
$\mathscr{N}_i$ has a maximum distance of $2^{s_i + 1}$ from $p_i$.

@@

\vspace*{-0.5em}
\begin{itemize}

@@

\end{itemize}
\vspace*{-0.5em}

@@

\vspace*{-0.8em}
\begin{equation}
| B_S(p, 2 \Delta) | \le c | B_S(p, \Delta) |\ \forall\ p \in S,\
\forall\ \Delta > 0.
\end{equation}
\vspace*{-1.4em}

@@

The expansion constant is used heavily in the cover tree literature.  It is
related to a notion of instrinic dimensionality, and in many scenarios, $c$ is
independent of the number of points in the dataset \cite{langford2006,
karger2002finding}.
%properties of cover trees in \cite{langford2006}.
One result we will use is that
the number of children of any cover tree node is $c^4$ (Lemma 4.1,
\cite{langford2006}), where $c$ is the expansion constant of the dataset the
cover tree is built on.

@@

Next we show the standard cover tree dual-tree pruning traversal, in a
problem-independent form.  This is given in Algorithm \ref{alg:cover-tree-dual}.
The algorithm is adapted from the original cover tree batch nearest neighbors
algorithm \cite{langford2006} and implements the dual-tree recursion for the
cover tree.  This traversal was first presented by Curtin and Ram
\cite{curtin2014dual}.
%
The correctness of this
traversal for various problem-dependent choices of the \texttt{BaseCase()} and
the \texttt{Score()} functions has been proven \cite{curtin2013tree}.
%  Now, we must introduce the inverse degree of bichromaticity.

\begin{algorithm}[htb]
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, set of reference nodes
$R$ \label{alg:line:ct-dual-input}
    \STATE {\bfseries Output:} none

    \medskip
    \STATE $s^{\max}_r \gets \max_{\mathscr{N}_r \in R} s_r$
    \IF{$(s_q < s^{\max}_r)$ {\bf or} $(s^{\max}_r = -\infty)$}
\label{alg:line:ct-dual-ref-recursion-start}
      \FORALL{$\mathscr{N}_r \in R$} \label{alg:line:ct-dual-base-case-start}
        \STATE \texttt{BaseCase($p_q$, $p_r$)}
      \ENDFOR \label{alg:line:ct-dual-base-case-end}
      \STATE $R_r \gets \{ \mathscr{N}_r \in R : s_r = s^{\max}_r \}$
\label{alg:line:ct-dual-ref-set}
      \STATE $R_{r - 1} \gets \{ \mathscr{C}(\mathscr{N}_r) : \mathscr{N}_r \in
R_r \} \cup (R \setminus R_r)$ \label{alg:line:ct-dual-ref-children}
      \STATE $R'_{r - 1} \gets \{ \mathscr{N}_r \in R_{r - 1} :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-ref-score}
      \STATE recurse with $\mathscr{N}_q$ and $R'_{r - 1}$
\label{alg:line:ct-dual-ref-recursion-end}
    \ELSE \label{alg:line:ct-dual-query-recursion-start}
      \FORALL{$\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q)$}
        \STATE $R' \gets \{ \mathscr{N}_r \in R :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-query-pruning}
        \STATE recurse with $\mathscr{N}_{qc}$ and $R'$
\label{alg:line:ct-dual-query-recursion}
      \ENDFOR \label{alg:line:ct-dual-query-recursion-end}
    \ENDIF
  \end{algorithmic}
  \caption{The standard pruning dual-tree traversal for cover trees.}
  \label{alg:cover-tree-dual}
\end{algorithm}

\begin{defn}
\label{def:bichromaticity}
Let $\mathscr{T}_q$ and $\mathscr{T}_r$ be two cover trees built on query set
$S_q$ and reference set $S_r$, respectively.  Now consider a pruning dual-tree
traversal (such as Algorithm \ref{alg:cover-tree-dual}) with the property that
the scales of nodes in $\mathscr{T}_q$ and $\mathscr{T}_r$ are kept as close as
possible---that is, the tree with the larger scale is always descended.  Then,
the {\bf inverse degree of bichromaticity} $\nu$ of the tree pair
$(\mathscr{T}_q, \mathscr{T}_r)$ is the maximum number of recursions in
$\mathscr{T}_r$ following a recursion in $\mathscr{T}_q$ before another
recursion in $\mathscr{T}_q$ or the termination of the algorithm (whichever
happens first).

This quantity is related to the {\bf degree of bichromaticity} $\kappa$
\cite{ram2009}, which is the maximum number of recursions in $\mathscr{T}_q$
between any two recursions in $\mathscr{T}_r$.
\end{defn}

Although $\kappa$ and $\nu$ can be difficult to bound, it is reasonable to
%monochromatic case where $S_q = S_r$, the trees $\mathscr{T}_q$ and
%$\mathscr{T}_r$ are identical, so $\kappa = \nu = 1$.  It is reasonable to
say that $\kappa \sim \nu \sim O(1)$ for datasets with roughly similar
distributions of pairwise distances.

@@

\vspace*{-0.7em}
\begin{equation}
| B_S(p, \rho \delta) \cap C | \le c^{2 + \log_2 \rho}.
\end{equation}
\vspace*{-1.1em}
\label{lem:packing}
\end{lemma}

\vspace*{-1em}
\begin{proof}
This is based on the packing argument from Lemma 4.1 in \cite{langford2006}.
Observe that $B_S(p, \rho \delta) \subseteq B_S(p_i, 2 \rho \delta)$ for any
$p_i \in S$, and that $| B_S(p, 2 \rho \delta) | = c^{2 + \log_2 \rho} |
B_S(p, \delta / 2) |$.  Because each point in $C$ is separated by $\delta$, the
number of points in $B_S(p, \rho \delta) \cap C$ is
bounded by the number of disjoint balls of radius $\delta / 2$ that can be
packed into $B_S(p, \rho \delta)$.  In the worst case, this packing is
perfect, and

\vspace*{-0.7em}
\begin{equation}
|B_S(p, \rho \delta)| \le \frac{|B_S(p_i, 2 \rho \delta)|}{|B_S(p_i, \delta
/ 2)|} \le c^{2 + \log_2 \rho}.
\end{equation}\end{proof}
\vspace*{-2.2em}

>>>>

As we will be making heavy use of trees, we must establish notation (taken from
\cite{curtin2013tree}).  The notation we will be using is defined in Table
\ref{tab:notation}.

\begin{table}
{\small
\begin{center}
\begin{tabular}{|c|l|}
\hline
{\bf Symbol} & {\bf Description} \\ \hline
$\mathscr{N}$ & A tree node \\ \hline
$\mathscr{C}_i$ & Set of child nodes of $\mathscr{N}_i$ \\ \hline
$\mathscr{P}_i$ & Set of points held in $\mathscr{N}_i$ \\ \hline
$\mathscr{D}_i^n$ & Set of descendant nodes of $\mathscr{N}_i$ \\ \hline
$\mathscr{D}_i^p$ & Set of points contained in $\mathscr{N}_i$ and
$\mathscr{D}_i^n$ \\ \hline
$\mu_i$ & Center of $\mathscr{N}_i$ (for cover trees, $\mu_i = p_i$) \\ \hline
$\lambda_i$ & Furthest descendant distance \\ \hline
\end{tabular}
\end{center}
}
\caption{Notation for trees.  See \cite{curtin2013tree} for details.}
\label{tab:notation}
\end{table}

The cover tree is a leveled hierarchical data structure originally proposed for
the task of nearest neighbor search by Beygelzimer
et~al.~\cite{langford2006}.  The original description of the tree describes both
an {\it implicit} and {\it explicit} representation, where the explicit
representation is the actual structure that would be implemented in practice.
Thus, we will consider only the explicit representation in this paper.

Each node $\mathscr{N}_i$ in the cover tree is indexed by an integer scale $s_i$
and holds one point $p_i$ which is also the center of the node ($p_i = \mu_i$).
The cover tree node may have a number of child nodes; if the cover tree node
does have children, then one child will be the self-child, which holds the same
point $p_i$ with a smaller scale $s_i$.  Each child of $\mathscr{N}_i$ will have
scale less than $s_i$, and every descendant point of a node $\mathscr{N}_i$
(that is, the set $\mathscr{D}_i^p$) has a maximum distance of $2^{s_i + 1}$
from the point $p_i$.

@@

\begin{itemize}

@@

\end{itemize}

@@

\begin{equation}
| B_S(p, 2 \Delta) | \le c | B_S(p, \Delta) |\ \forall\ p \in S,\
\forall\ \Delta > 0.
\end{equation}

@@

The expansion constant is used heavily in the cover tree literature.  It is
related to a notion of instrinic dimensionality, and in many scenarios, $c$ is
independent of the number of points in the dataset \cite{langford2006,
karger2002finding}.

The expansion constant can be used to show a few useful bounds on various
properties of the cover tree; we restate these results below, given some cover
tree built on a dataset $S$ with expansion constant $c$ and $|S| = N$:

\begin{itemize}
  \item {\bf Width bound:} no cover tree node has more than $c^4$ children
(Lemma 4.1, \cite{langford2006}).

  \item {\bf Depth bound:} the maximum depth of any node is $O(c^2 \log N)$
(Lemma 4.3, \cite{langford2006}).

  \item {\bf Space bound:} a cover tree has $O(N)$ nodes (Theorem 1,
\cite{langford2006}).
\end{itemize}

@@

@@

\begin{equation}
| B_S(p, \rho \delta) \cap C | \le c^{2 + \log_2 \rho}.
\end{equation}
\label{lem:packing}
\end{lemma}

\begin{proof}
This is based on the packing argument from Lemma 4.1 in \cite{langford2006}.
Observe that $B_S(p, \rho \delta) \subseteq B_S(p_i, 2 \rho \delta)$ for any
$p_i \in S$, and that $| B_S(p, 2 \rho \delta) | = c^{2 + \log_2 \rho} |
B_S(p, \delta / 2) |$.  Because each point in $C$ is separated by $\delta$, the
number of points in $B_S(p, \rho \delta) \cap C$ is
bounded by the number of disjoint balls of radius $\delta / 2$ that can be
packed into $B_S(p, \rho \delta)$.  In the worst case, this packing is
perfect, and

\begin{equation}
|B_S(p, \rho \delta)| \le \frac{|B_S(p_i, 2 \rho \delta)|}{|B_S(p_i, \delta
/ 2)|} \le c^{2 + \log_2 \rho}.
\end{equation}\end{proof}
